---
title: "Student Performance Analysis Code"
author: "Katie Romrell"
date: "2025-06-10"
output: html_document
---
NOTE: All datasets have been excluded to preserve confidentiality. Instead, this code includes dummy data that was generated by ChatGPT to make the code executable but will not show the actual trends found in this analysis.

## Set up Environment

Loading packages needed for this analysis.

```{r Environment, include=FALSE}
library(tidyverse)  
library(scales)     # For formatting percentages
library(patchwork)  # For multi-plot layouts
library(relaimpo)   # For relative importance
library(broom)      # Used in support group slope analysis
library(purrr)      # To run models across support groups
```

Fake data tibble to make code executable.

```{r}
```{r Create_Fake_Data}

all_student_data <- tibble(
  grade_point_dec = round(runif(200, 0, 4), 2),
  absence_rate = sample(c("low", "medium", "high", "very high"), 200, replace = TRUE),
  absences = sample(0:15, 200, replace = TRUE),       # Now 0–15
  tardies = sample(0:15, 200, replace = TRUE),         # Now 0–15
  ss_absences = sample(0:15, 200, replace = TRUE),     # Now 0–15
  ell = sample(0:1, 200, replace = TRUE),
  sped = sample(0:1, 200, replace = TRUE),
  sec_504 = sample(0:1, 200, replace = TRUE),
  tag = sample(0:1, 200, replace = TRUE),
  pass_or_fail = sample(c(0, 1), 200, replace = TRUE)
)

all_with_core <- all_student_data %>%
  mutate(core_req = sample(c(0, 1), n(), replace = TRUE))


```

## Finding 1: Absences predict lower student success

### Dashboard 1: Absences Impact on Grades/Pass Rates

Summarizing grade point and pass rate by absence category

```{r Absences vs. grade/pass rates}
grade_by_absence <- all_student_data %>%
  group_by(absence_rate) %>%
  summarise(
    mean_grade_point = mean(grade_point_dec, na.rm = TRUE),
    count = n()
  ) %>%
  mutate(
    absence_rate_desc = factor(
      absence_rate,
      levels = c("low", "medium", "high", "very high"),
      labels = c(
        "Low (0-9%)",
        "Medium (10-19%)",
        "High (20-39%)",
        "Very High (40%+)"
      )
    )
  )

pass_by_absence <- all_student_data %>%
  group_by(absence_rate) %>%
  summarise(
    pass_rate = mean(pass_or_fail),
    count = n()
  ) %>%
  mutate(
    absence_rate_desc = factor(
      absence_rate,
      levels = c("low", "medium", "high", "very high"),
      labels = c(
        "Low (0-9%)",
        "Medium (10-19%)",
        "High (20-39%)",
        "Very High (40%+)"
      )
    )
  )
```

Plotting Absences vs. Grades/Pass Rates

```{r Plotting Absence vs. Grades/Pass Rates}

# Plot 1: Average Grade by Absences
p1 <- ggplot(grade_by_absence, aes(x = absence_rate_desc,
                                   y = mean_grade_point,
                                   fill = absence_rate_desc)) +
  geom_col() +
  geom_text(aes(label = round(mean_grade_point, 2)), vjust = -0.5, size = 4) +
  geom_text(aes(label = paste0("n=", count)), vjust = 1.5, color = "white", size = 3) +
  scale_fill_manual(values = c(
    "Low (0-9%)" = "#3CB371",
    "Medium (10-19%)" = "gold",
    "High (20-39%)" = "orange",
    "Very High (40%+)" = "red"
  )) +
  labs(
    title = "Avg Grade by Absences",
    x = "Absence Rate",
    y = "Average Grade Point"
  ) +
  coord_cartesian(ylim = c(0, 4)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(margin = margin(t = 5))
  )

# Plot 2: Pass Rate by Absences
p2 <- ggplot(pass_by_absence, aes(x = absence_rate_desc,
                                  y = pass_rate,
                                  fill = absence_rate_desc)) +
  geom_col() +
  geom_text(aes(label = scales::percent(pass_rate, accuracy = 1)), vjust = -0.5, size = 4) +
  geom_text(aes(label = paste0("n=", count)), vjust = 1.5, color = "white", size = 3) +
  scale_fill_manual(values = c(
    "Low (0-9%)" = "#3CB371",
    "Medium (10-19%)" = "gold",
    "High (20-39%)" = "orange",
    "Very High (40%+)" = "red"
  )) +
  labs(
    title = "Pass Rate by Absences",
    x = "Absence Rate",
    y = "Pass Rate"
  ) +
  coord_cartesian(ylim = c(0, 1.05)) +  # extended to 105%
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(margin = margin(t = 5))
  )

# Combine plots into a 2x1 grid
combined_dashboard_reg <- (p1 | p2) + 
  plot_annotation(
    title = "Absences Impact on Grades/Pass Rates (by Class)",
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
  )

# Print combined dashboard
print(combined_dashboard_reg)


```

## Table 1: Factors Explaining Grade Point

Calculating relative importance of each variables in the multiple regression model.
```{r}
gpa_model_all <- lm(grade_point_dec ~ absences + tardies + ss_absences + ell + sped + sec_504 + tag, data = all_student_data)

summary(gpa_model_all)

calc.relimp(gpa_model_all, type = "lmg", rela = TRUE)
```
## Finding 2: Core classes and non-core classes vary by predictor/success rates

### Dashboard 2:  Absences Impact on Grades/Pass Rates (by Class Type)

Separate and group the data
```{r}

# Separate core vs. non-core classes
core_data <- all_with_core %>% filter(core_req == 1)
non_core_data <- all_with_core %>% filter(core_req == 0)

# Group by core/non-core and GPA/pass or fail

grade_by_absence_core <- core_data %>%
  group_by(absence_rate) %>%
  summarise(mean_grade_point = mean(grade_point_dec, na.rm = TRUE), count = n()) %>%
  mutate(absence_rate_desc = factor(absence_rate,
           levels = c("low", "medium", "high", "very high"),
           labels = c("Low (0-9%)", "Medium (10-19%)", "High (20-39%)", "Very High (40%+)")))

pass_by_absence_core <- core_data %>%
  group_by(absence_rate) %>%
  summarise(pass_rate = mean(pass_or_fail), count = n()) %>%
  mutate(absence_rate_desc = factor(absence_rate,
           levels = c("low", "medium", "high", "very high"),
           labels = c("Low (0-9%)", "Medium (10-19%)", "High (20-39%)", "Very High (40%+)")))

grade_by_absence_non_core <- non_core_data %>%
  group_by(absence_rate) %>%
  summarise(mean_grade_point = mean(grade_point_dec, na.rm = TRUE), count = n()) %>%
  mutate(absence_rate_desc = factor(absence_rate,
           levels = c("low", "medium", "high", "very high"),
           labels = c("Low (0-9%)", "Medium (10-19%)", "High (20-39%)", "Very High (40%+)")))

pass_by_absence_non_core <- non_core_data %>%
  group_by(absence_rate) %>%
  summarise(pass_rate = mean(pass_or_fail), count = n()) %>%
  mutate(absence_rate_desc = factor(absence_rate,
           levels = c("low", "medium", "high", "very high"),
           labels = c("Low (0-9%)", "Medium (10-19%)", "High (20-39%)", "Very High (40%+)")))
```

Create dashboard with four graphs, demonstrating the correlation of absences on GPA/pass rates as separated by core vs. non-core

```{r}

# Plot 3: Average Grade by Absences (Core)
p3 <- ggplot(grade_by_absence_core, aes(x = absence_rate_desc,
                                        y = mean_grade_point,
                                        fill = absence_rate_desc)) +
  geom_col() +
  geom_text(aes(label = round(mean_grade_point, 2)), vjust = -0.5, size = 4) +
  geom_text(aes(label = paste0("n=", count)), vjust = 1.5, color = "white", size = 3) +
  scale_fill_manual(values = c(
    "Low (0-9%)" = "#3CB371",
    "Medium (10-19%)" = "gold",
    "High (20-39%)" = "orange",
    "Very High (40%+)" = "red"
  )) +
  labs(
    title = "Avg Grade by Absences (Core)",
    x = "Absence Rate",
    y = "Average Grade Point"
  ) +
  coord_cartesian(ylim = c(0, 4)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(margin = margin(t = 5))
  )

# Plot 4: Pass Rate by Absences (Core Classes)
p4 <- ggplot(pass_by_absence_core, aes(x = absence_rate_desc,
                                       y = pass_rate,
                                       fill = absence_rate_desc)) +
  geom_col() +
  geom_text(aes(label = scales::percent(pass_rate, accuracy = 1)), vjust = -0.5, size = 4) +
  geom_text(aes(label = paste0("n=", count)), vjust = 1.5, color = "white", size = 3) +
  scale_fill_manual(values = c(
    "Low (0-9%)" = "#3CB371",
    "Medium (10-19%)" = "gold",
    "High (20-39%)" = "orange",
    "Very High (40%+)" = "red"
  )) +
  labs(
    title = "Pass Rate by Absences (Core)",
    x = "Absence Rate",
    y = "Pass Rate"
  ) +
  coord_cartesian(ylim = c(0, 1.05)) +  # extended to 105%
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(margin = margin(t = 5))
  )

# Plot 5: Average Grade by Absences (non-core)
p5 <- ggplot(grade_by_absence_non_core, aes(x = absence_rate_desc,
                                   y = mean_grade_point,
                                   fill = absence_rate_desc)) +
  geom_col() +
  geom_text(aes(label = round(mean_grade_point, 2)), vjust = -0.5, size = 4) +
  geom_text(aes(label = paste0("n=", count)), vjust = 1.5, color = "white", size = 3) +
  scale_fill_manual(values = c(
    "Low (0-9%)" = "#3CB371",
    "Medium (10-19%)" = "gold",
    "High (20-39%)" = "orange",
    "Very High (40%+)" = "red"
  )) +
  labs(
    title = "Avg Grade by Absences (Non-Core)",
    x = "Absence Rate",
    y = "Average Grade Point"
  ) +
  coord_cartesian(ylim = c(0, 4)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(margin = margin(t = 5))
  )

# Plot 6: Pass Rate by Absences (non-core)
p6 <- ggplot(pass_by_absence_non_core, aes(x = absence_rate_desc,
                                  y = pass_rate,
                                  fill = absence_rate_desc)) +
  geom_col() +
  geom_text(aes(label = scales::percent(pass_rate, accuracy = 1)), vjust = -0.5, size = 4) +
  geom_text(aes(label = paste0("n=", count)), vjust = 1.5, color = "white", size = 3) +
  scale_fill_manual(values = c(
    "Low (0-9%)" = "#3CB371",
    "Medium (10-19%)" = "gold",
    "High (20-39%)" = "orange",
    "Very High (40%+)" = "red"
  )) +
  labs(
    title = "Pass Rate by Absences (Non-Core)",
    x = "Absence Rate",
    y = "Pass Rate"
  ) +
  coord_cartesian(ylim = c(0, 1.05)) +  # extended to 105%
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(margin = margin(t = 5))
  )

# Combine chosen four plots (non-core and core) into a 2x2 grid
combined_dashboard_type <- (p3 | p4) / (p5 | p6) + 
  plot_annotation(
    title = "Absences Impact on Grades/Pass Rates (by Class Type)",
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
  )

# Print combined dashboard
print(combined_dashboard_type)
```

### Table 2: Factors Explaining Grade Point (by Course Type)

```{r}
gpa_model_core <- lm(grade_point_dec ~ absences + tardies + ell + sped + sec_504 + tag + ss_absences, data = core_data)
summary(gpa_model_core)
calc.relimp(gpa_model_core, type = "lmg", rela = TRUE)

gpa_model_non_core <- lm(grade_point_dec ~ absences + tardies + ell + sped + sec_504 + tag + ss_absences, data = non_core_data)
summary(gpa_model_non_core)
calc.relimp(gpa_model_non_core, type = "lmg", rela = TRUE)
```

## Finding 3: Support needs explains decrease in grades; TAG explains increase in grades

### Graph 3: Grade Point vs. Absences by Support Status (Core Classes)

```{r}
# Translate four binary support group variables into one categorical variable
core_data_with_support <- all_with_core %>%
  mutate(support_group = case_when(
    ell == 1 ~ "ELL",
    sped == 1 ~ "SPED",
    sec_504 == 1 ~ "504",
    tag == 1 ~ "TAG",
    TRUE ~ "None"
  ))

# Calculate slope of GPA ~ absences for each group
core_slope_labels <- core_data_with_support %>%
  group_by(support_group) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(grade_point_dec ~ absences, data = .x)),
    tidied = map(model, tidy)
  ) %>%
  unnest(tidied) %>%
  filter(term == "absences") %>%
  mutate(slope = estimate) %>%
  { .[, c("support_group", "slope")] }

# Choose label position
core_label_positions <- core_data_with_support %>%
  group_by(support_group) %>%
  summarize(x = max(absences, na.rm = TRUE) * 0.7, y = min(grade_point_dec, na.rm = TRUE) + 0.3)

# Merge slope values with label positions
core_slope_labels <- left_join(core_slope_labels, core_label_positions, by = "support_group")

# Plot the data
ggplot(core_data_with_support, aes(x = absences, y = grade_point_dec, color = support_group)) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(
    data = core_slope_labels,
    aes(x = x, y = y, label = paste0("slope = ", round(slope, 3))),
    inherit.aes = FALSE,
    color = "grey30",
    size = 3,
    vjust = -0.5
  ) +
  facet_wrap(~ support_group) +
  scale_y_continuous() +  # optional
  labs(
    title = "Grade Point vs. Absences by Support Status (Core Classes)",
    x = "Absences for Class",
    y = "Grade Point for Class"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

### Graph 4: Grade Point vs. Absences by Support Status (Non-core Classes)

```{r}
# Translate four binary support group variables into one categorical variable
non_core_data_with_support <- non_core_data %>%
  mutate(
    support_group = case_when(
      ell == 1 ~ "ELL",
      sped == 1 ~ "SPED",
      sec_504 == 1 ~ "504",
      tag == 1 ~ "TAG",
      TRUE ~ "None"
    )
  )

non_core_slope_labels <- non_core_data_with_support %>%
  group_by(support_group) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(grade_point_dec ~ absences, data = .x)),
    tidied = map(model, tidy)
  ) %>%
  unnest(tidied) %>%
  filter(term == "absences") %>%
  mutate(slope = estimate) %>%
  { .[, c("support_group", "slope")] }

# Choose label position
non_core_label_positions <- non_core_data_with_support %>%
  group_by(support_group) %>%
  summarize(x = max(absences, na.rm = TRUE) * 0.7, y = min(grade_point_dec, na.rm = TRUE) + 0.3)

# Merge slope values with label positions
non_core_slope_labels <- left_join(non_core_slope_labels, non_core_label_positions, by = "support_group")

# Plot the data
ggplot(non_core_data_with_support, aes(x = absences, y = grade_point_dec, color = support_group)) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(
    data = non_core_slope_labels,
    aes(x = x, y = y, label = paste0("slope = ", round(slope, 3))),
    inherit.aes = FALSE,
    color = "grey30",
    size = 3,
    vjust = -0.5
  ) +
  facet_wrap(~ support_group) +
  scale_y_continuous() +  # optional
  labs(
    title = "Grade Point vs. Absences by Support Status (Non-core Classes)",
    x = "Absences for Class",
    y = "Grade Point for Class"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

